`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities

### 2020-07-19 ms.Rmd
Added citation for cor_phylo covariate method.
update ms and ref

1. Model-based approaches are increasingly popular in ecological studies. A good example of this trend is the use of joint species distribution models to ask questions about ecological communities. However, most current applications of model-based methods do not include phylogenies despite the well-known importance of phylogenetic relationships in shaping species distributions and community composition. In part, this is due to a lack of accessible tools allowing ecologists to fit phylogenetic species distribution models easily.
2. To fill this gap, the R package `phyr` (pronounced fire) implements a suite of metrics, comparative methods and mixed models that use phylogenies to understand and predict community composition and other ecological and evolutionary phenomena. The `phyr` workhorse functions are implemented in C++ making all calculations and model estimations fast.
3. `phyr` can fit a variety of models such as phylogenetic joint-species distribution models, spatiotemporal-phylogenetic autocorrelation models, and phylogenetic trait‐based bipartite network models. `phyr` also estimates phylogenetically independent trait correlations with measurement error to test for adaptive syndromes and performs fast calculations of common alpha and beta phylogenetic diversity metrics. All `phyr` methods are united under Brownian motion or Ornstein-Uhlenbeck models of evolution and phylogenetic terms are modelled as phylogenetic covariance matrices.
4. The functions and model formula syntax we propose in `phyr` serves as a simple and unified framework that ignites the use of phylogenies to address a variety of ecological questions.

@@

Here, Y is a binary (Bernoulli) dependent variable which takes values of either 0 or 1. The specification `family = 'binomial'` allows binary data and also binomial data for which Y is a matrix containing columns for successes and failures. The independent variables `trait` and `env` take on different values for each species and site, respectively. Sites (`site`) and species (`sp`) are treated as random effects: `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites. For the case of species, the double underscore in `(1|sp__)` implies that, in addition to a random effect for species, there is a second random effect which contains the phylogenetic relationships among species (or some other correlation structure specified by the user). The phylogenetic random effect assumes that values for each species are picked from a multivariate Gaussian distribution with phylogenetic covariance matrix $\Sigma$. A covariance matrix $\Sigma$ is specified by `cov_ranef = list(sp = phy.sp, site = V.space)`. The covariance matrix `phy.sp` associated with species can be a `phylo` object from the R package `ape` [@ape2018]. To construct $\Sigma$ from a 'phylo' object, `pglmm()` assumes that the residual variation associated with species follows a Brownian motion model of evolution, so that the covariance between species is proportional to their shared evolutionary history (e.g., shared branch length on a phylogeny). It is also possible to specify an explicit covariance matrix, such as `site = V.space`, where `V.space` is a covariance matrix created from the distance between sites.

@@

Earlier versions of `pglmm()` and `cor_phylo()` both appear in existing R packages (`pez` [@pearse2015pez] and `ape` [@ape2018], respectively), although the versions in `phyr` represent considerable improvements in ease-of-use, computational speed, and flexibility. Both have new syntax that makes them more intuitive to use. `pglmm()` also has new associated functions that plot the design of the covariance matrices (Fig. \@ref(fig:designPlot)), making model interpretation easier. Both are now coded in C++ (for key functions), which speeds computation time by 5-10X. `pglmm()` now supports several non-Gaussian distributions and allows Bayesian analyses using INLA [@rue2009approximate] that is particularly useful for large datasets. Finally, both include more output; for example, both now include facilities to perform likelihood ratio tests and compute AIC and BIC values for model comparisons.

@@

`pglmm()` is syntactically modeled after `lmer()` and `glmer()` in `lme4` [@bates2015lme4], although it allows the specification of phylogenetic covariance matrices. `pglmm()` also allows "nested" models (with block-diagonal covariance matrices) which arise when phylogenetic covariances only act within single communities, rather than among communities; an example is illustrated by the `(1|sp__@site)` term in Fig. \@ref(fig:designPlot). Such nested models make it possible to assess whether phylogenetic relatedness affects the abundance of species within the same communities, such as whether competition between closely related species excludes one of the competitors from communities where the other is present. Nested models are structurally incompatible with the architecture of `lme4`.

@@

We fitted a PGLMM that examined how a hypothetical functional trait, environmental gradient, and their interaction affect distributions of `r nspp` species across `r nsite` sites. We focused on abundance and used the default family of data distribution (Gaussian), but other distributions can also be specified by resetting the `family` argument. Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(sp = phy, site = V.space)` where `sp` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a covariance matrix among sites. This model can also be fitted with Bayesian framework by setting `bayes = TRUE`, which is recommended when data set is large.

>>>>

1. Model-based approaches are increasingly popular in ecological studies. A good example of this trend is the use of joint species distribution models to ask questions about ecological communities. However, most current applications of model-based methods do not include phylogenies despite the well-known importance of phylogenetic relationships in shaping species distributions and community composition. In part, this is due to a lack of accessible tools allowing ecologists to fit phylogenetic species distribution models easily.
2. To fill this gap, the R package `phyr` (pronounced fire) implements a suite of metrics, comparative methods and mixed models that use phylogenies to understand and predict community composition and other ecological and evolutionary phenomena. The `phyr` workhorse functions are implemented in C++ making all calculations and model estimations fast.
3. `phyr` can fit a variety of models such as phylogenetic joint-species distribution models, spatiotemporal-phylogenetic autocorrelation models, and phylogenetic trait‐based bipartite network models. `phyr` also estimates phylogenetically independent trait correlations with measurement error to test for adaptive syndromes and performs fast calculations of common alpha and beta phylogenetic diversity metrics. All `phyr` methods are united under Brownian motion or Ornstein-Uhlenbeck models of evolution and phylogenetic terms are modelled as phylogenetic covariance matrices.
4. The functions and model formula syntax we propose in `phyr` provide an easy-to-use collection of tools that we hope will ignite the use of phylogenies to address a variety of ecological questions.

@@

Here, Y is a binary (Bernoulli) dependent variable which takes values of either 0 or 1. The specification `family = 'binomial'` allows binary data and also binomial data for which Y is a matrix containing columns for successes and failures. The independent variables `trait` and `env` take on different values for each species and site, respectively. Sites (`site`) and species (`sp`) are treated as random effects: `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites. For the case of species, the double underscore in `(1|sp__)` implies that, in addition to a random effect for species, there is a second random effect which contains the phylogenetic relationships among species (or some other correlation structure specified by the user). The phylogenetic random effect assumes that values for each species are picked from a multivariate Gaussian distribution with phylogenetic covariance matrix $\Sigma$. A covariance matrix $\Sigma$ is specified by `cov_ranef = list(sp = phy.sp, site = V.space)`. The covariance matrix `phy.sp` associated with species can be a `phylo` object from the R package `ape` [@ape2018]. To construct $\Sigma$ from a 'phylo' object, `pglmm()` assumes that the residual variation associated with species follows a Brownian motion model of evolution, so that the covariance between species is proportional to their shared evolutionary history (e.g., shared branch length on a phylogeny). It is also possible to specify an explicit covariance matrix, such as `site = V.space`, where `V.space` is a covariance matrix created from the distance between sites. For example, if we assume that spatial correlations follow a Gaussian function, then the correlation in residuals from sites _i_ and _j_ located a distance $d_{ij}$ from each other is $\exp(-(d_{ij}/r)^2)$, where _r_ is the "range" giving how quickly spatial correlation decreases with distance [@besag1975estimation].

@@

Earlier versions of `pglmm()` and `cor_phylo()` both appear in existing R packages (`pez` [@pearse2015pez] and `ape` [@ape2018], respectively), although the versions in `phyr` represent considerable improvements in ease-of-use, computational speed, and flexibility. Both have new syntax that makes them more intuitive to use. `pglmm()` also has new associated functions that plot the design of the covariance matrices (Fig. \@ref(fig:designPlot)), making model interpretation easier. Both are now coded in C++ (for key functions), which speeds computation time by 5-10X. `pglmm()` now supports several non-Gaussian distributions and allows Bayesian analyses using INLA [@rue2009approximate] that is particularly useful for large datasets. Finally, both include more output; for example, both now report the log-likelihood and related measures AIC and BIC. The log-likelihood makes it possible to compare full models to reduced models with some parameters removed, using likelihood ratio test [e.g., @neter1989applied] to give the significance of a collection of parameters. AIC and BIC can be used for model selection [@burnham2002model].

@@

`pglmm()` is syntactically modeled after `lmer()` and `glmer()` in `lme4` [@bates2015lme4], although it allows the specification of phylogenetic covariance matrices. `pglmm()` also allows "nested" models (with block-diagonal covariance matrices) which arise when phylogenetic covariances only act within single communities, rather than among communities; an example is illustrated by the `(1|sp__@site)` term in Fig. \@ref(fig:designPlot). Such nested models make it possible to assess whether phylogenetic relatedness affects the abundance of species within the same communities, such as whether competition between closely related species excludes one of the competitors from communities where the other is present. This is the case in the distribution of fish among lakes: after accounting for the effect of pH excluding some species from some lakes, a pattern of exclusion occurs in which phylogenetically related species are less likely to occur in the same lakes [@helmus2010communities]. Nested models are structurally incompatible with the architecture of `lme4`.

@@

We fitted a PGLMM that examined how a hypothetical functional trait, environmental gradient, and their interaction affect distributions of `r nspp` species across `r nsite` sites. We focused on abundance and used the default family of data distribution (Gaussian), but other distributions can also be specified by resetting the `family` argument. Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(sp = phy, site = V.space)` where `sp` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `V.space` is a covariance matrix among sites. This model can also be fitted with Bayesian framework by setting `bayes = TRUE`, which is recommended when data set is large.
